---
title: "Lab 4: Spatial Predictive Analysis"
subtitle: "Analyzing 311 Violations"
author: "Huda Abdelnur"
date: today
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

## Data Loading & Exploration

```{r, echo=FALSE}
##library in necessary packages 
library(tidyverse)      # Data manipulation
library(sf)             # Spatial operations
library(here)           # Relative file paths
library(viridis)        # Color scales
library(terra)          # Raster operations (replaces 'raster')
library(spdep)          # Spatial dependence
library(FNN)            # Fast nearest neighbors
library(MASS)           # Negative binomial regression
library(patchwork)      # Plot composition (replaces grid/gridExtra)
library(knitr)          # Tables
library(kableExtra)     # Table formatting
library(classInt)   
library(ggspatial)

# Spatstat for analyzing point patterns - split into sub-packages
library(spatstat.geom)    # Spatial geometries
library(spatstat.explore) # Spatial exploration/KDE

# Set options
options(scipen = 999)  # No scientific notation
set.seed(5080)         # Reproducibility, could be any number

```


```{r, results='hide'}
##Loading in 311 data and Chicago geographic boundaries 
chicagoBoundary <- 
  st_read("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/DATA/Chapter5/chicagoBoundary.geojson") %>%
  st_transform('ESRI:102271')

vacant <- read_csv("C:/Users/hudaa/Documents/CPLN5920/hudabdel-sp-cpln5920/labs/lab_4/data/311_Service_Requests_-_Vacant_and_Abandoned_Buildings_Reported_-_Historical_20260409.csv")

burglaries <- st_read("C:/Users/hudaa/Downloads/burglaries.shp") %>% 
  st_transform('ESRI:102271')
```

```{r}
##clean data to keep cases from 2015 and beyond. Data only has up until 2018
vacant$`DATE SERVICE REQUEST WAS RECEIVED` <- as.Date(vacant$`DATE SERVICE REQUEST WAS RECEIVED`, format = "%m/%d/%Y")
vacant <- vacant[format(vacant$`DATE SERVICE REQUEST WAS RECEIVED`, "%Y") == "2017", ]


##transform vacant csv to spatial dataframe
vacant <- vacant %>%
  drop_na("X COORDINATE", "Y COORDINATE") %>%
  st_as_sf(coords = c("X COORDINATE", "Y COORDINATE"), crs = 3435) %>%  # Illinois State Plane feet
  st_transform(crs = "ESRI:102271")
```

```{r}
##creating a theme to make maps consistent 

theme_crime <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 1),
      plot.subtitle = element_text(color = "gray30", size = base_size - 1),
      legend.position = "right",
      panel.grid.minor = element_blank(),
      axis.text = element_blank(),
      axis.title = element_blank()
    )
}

# Set as default
theme_set(theme_crime())

##point map of vacant building locations

p1 <- ggplot() + 
  geom_sf(data = chicagoBoundary, fill = "gray95", color = "gray60") +
  geom_sf(data = vacant, color = "#d62828", size = 0.1, alpha = 0.4) +
  annotation_north_arrow(
    location = "bl",          # bottom left — or "br", "tl", "tr"
    which_north = "true",
    style = north_arrow_fancy_orienteering(),  # or north_arrow_minimal()
    pad_x = unit(0.2, "in"),
    pad_y = unit(0.2, "in")
  ) +
  labs(
    title = "Vacant Building Locations",
    subtitle = paste0("Chicago 2015-2018, n = ", nrow(vacant))
  ) +
  theme_void() +
  theme(plot.title = element_text(size = 10),
        plot.subtitle = element_text(size = 8))

##density map of vacant buildings
p2 <- ggplot() + 
  geom_sf(data = chicagoBoundary, fill = "gray95", color = "gray60") +
  geom_density_2d_filled(
    data = data.frame(st_coordinates(vacant)),
    aes(X, Y),
    alpha = 0.7,
    bins = 8
  ) +
  coord_sf() +  # explicitly force coord_sf to align with chicagoBoundary
  scale_fill_viridis_d(
    option = "plasma",
    direction = -1,
    guide = "none"  
  ) +
  labs(
    title = "Density Surface",
    subtitle = "Kernel density estimation"
  ) +
  theme_void() +
  theme(plot.title = element_text(size = 10),
        plot.subtitle = element_text(size = 8))

p1 + p2 + 
  plot_annotation(
    title = "Spatial Distribution of Vacant and Abandoned Buildings in Chicago",
    tag_levels = "A",
    theme = theme(plot.title = element_text(size = 12, hjust = 0.5))
  )
```

## Fishnet Grid Creation

```{r}
##create a 500M X 500M fishnet
fishnet <- st_make_grid(
  chicagoBoundary,
  cellsize = 500,  # 500 meters per cell
  square = TRUE
) %>%
  st_sf() %>%
  mutate(uniqueID = row_number()) %>%
  st_transform('ESRI:102271') 

# Keep only cells that intersect Chicago
fishnet <- fishnet[chicagoBoundary, ]

# Spatial join: which cell contains each burglary?
burglaries_fishnet <- st_join(burglaries, fishnet, join = st_within) %>%
  st_drop_geometry() %>%
  group_by(uniqueID) %>%
  summarize(countBurglaries = n(), .groups = "drop")


# Join back to fishnet (cells with 0 burglaries will be NA)
fishnet <- fishnet %>%
  left_join(burglaries_fishnet, by = "uniqueID") %>%
  mutate(countBurglaries = replace_na(countBurglaries, 0))


# Spatial join: which cells have vacant properties?
vacant_fishnet <- st_join(vacant, fishnet, join = st_within) %>%
  st_drop_geometry() %>%
  group_by(uniqueID) %>%
  summarize(countvacant = n()) 

# Join back to fishnet (cells with 0 vacant buildings will be NA)
fishnet <- fishnet %>%
  left_join(vacant_fishnet, by = "uniqueID") %>%
  mutate(countvacant = replace_na(countvacant, 0))

# Summary statistics

cat("\nCells with zero vacant buildings:", 
    sum(fishnet$countvacant == 0), 
    "/", nrow(fishnet),
    "(", round(100 * sum(fishnet$countvacant == 0) / nrow(fishnet), 1), "%)\n")
```

```{r visualize-fishnet}
#| fig-width: 8
#| fig-height: 6

# Visualize aggregated counts
p3 <- ggplot() +
  geom_sf(data = fishnet, aes(fill = countvacant), color = NA) +
  geom_sf(data = chicagoBoundary, fill = NA, color = "white", linewidth = 1) +
  scale_fill_viridis_c(
    name = "Vacant Buildigns",
    option = "plasma",
    trans = "sqrt",  # Square root for better visualization of skewed data
    breaks = c(0, 1, 5, 10, 20, 40)
  ) +
  labs(
    title = "Vacant Building Counts by Grid Cell",
    subtitle = "500m x 500m cells, Chicago 2017"
  ) +
  theme_crime()

p4 <- ggplot() +
  geom_sf(data = fishnet, aes(fill = countBurglaries), color = NA) +
  geom_sf(data = chicagoBoundary, fill = NA, color = "white", linewidth = 1) +
  scale_fill_viridis_c(
    name = "Burglaries",
    option = "plasma",
    trans = "sqrt",  # Square root for better visualization of skewed data
    breaks = c(0, 1, 5, 10, 20, 40)
  ) +
  labs(
    title = "Burglary Counts by Grid Cell",
    subtitle = "500m x 500m cells, Chicago 2017"
  ) +
  theme_crime()

p3 + p4 + plot_annotation(
    title = "Spatial Distribution of Burglaries and Vacant Buildings in Chicago",
    tag_levels = "A",
    theme = theme(plot.title = element_text(size = 12, hjust = 0.5))
  )



```

```{r kde-baseline}
#| message: false

# Convert vacant to point pattern analysis. 
vacant_ppp <- as.ppp(
  st_coordinates(vacant),
  W = as.owin(st_bbox(chicagoBoundary))
)

# Calculate KDE with 1km bandwidth
kde_vacant <- density.ppp(
  vacant_ppp,
  sigma = 1000,  # 1km bandwidth
  edge = TRUE    # Edge correction
)

# Convert to terra raster (modern approach, not raster::raster)
kde_raster <- rast(kde_vacant)

# Extract KDE values to fishnet cells
fishnet <- fishnet %>%
  mutate(
    kde_value = terra::extract(
      kde_raster,
      vect(fishnet),
      fun = mean,
      na.rm = TRUE
    )[, 2]  # Extract just the values column
  )


# Convert vacant to point pattern analysis. 
b_ppp <- as.ppp(
  st_coordinates(burglaries),
  W = as.owin(st_bbox(chicagoBoundary))
)

# Calculate KDE with 1km bandwidth
kde_b <- density.ppp(
  b_ppp,
  sigma = 1000,  # 1km bandwidth
  edge = TRUE    # Edge correction
)

# Convert to terra raster (modern approach, not raster::raster)
kde_raster_b <- rast(kde_b)

# Extract KDE values to fishnet cells
fishnet <- fishnet %>%
  mutate(
    kde_b = terra::extract(
      kde_raster_b,
      vect(fishnet),
      fun = mean,
      na.rm = TRUE
    )[, 2]  # Extract just the values column
  )

```

```{r visualize-kde}
#| fig-width: 8
#| fig-height: 6

p5 <- ggplot() +
  geom_sf(data = fishnet, aes(fill = kde_value), color = NA) +
  geom_sf(data = chicagoBoundary, fill = NA, color = "white", linewidth = 1) +
  scale_fill_viridis_c(
    name = "KDE Value",
    option = "plasma"
  ) +
  labs(
    title = "Kernel Density Estimation Baseline",
    subtitle = "Simple spatial smoothing of vacant building locations"
  ) +
  theme_crime()


p6 <- ggplot() +
  geom_sf(data = fishnet, aes(fill = kde_b), color = NA) +
  geom_sf(data = chicagoBoundary, fill = NA, color = "white", linewidth = 1) +
  scale_fill_viridis_c(
    name = "KDE Value",
    option = "plasma"
  ) +
  labs(
    title = "Kernel Density Estimation Baseline",
    subtitle = "Simple spatial smoothing of burglary locations"
  ) +
  theme_crime()

p5 + p6 + plot_annotation(
    title = "Kernely Density Maps of Burglaries and Vacant Buildings in Chicago",
    tag_levels = "A",
    theme = theme(plot.title = element_text(size = 12, hjust = 0.5))
  )
```


## Spatial Features

```{r nn-feature}
#| message: false

# Calculate mean distance to 3 nearest vacant buildings
# (Do this OUTSIDE of mutate to avoid sf conflicts)

# Get coordinates
fishnet_coords <- st_coordinates(st_centroid(fishnet))
vacant_coords <- st_coordinates(vacant)

# Calculate k nearest neighbors and distances.
nn_result <- get.knnx(vacant_coords, fishnet_coords, k = 3)

# Add to fishnet
fishnet <- fishnet %>%
  mutate(
    vacant.nn = rowMeans(nn_result$nn.dist)
  )

summary(fishnet$vacant.nn)
```

```{r local-morans-vacant buildings}
# Function to calculate Local Moran's I, essentially this function will be applied to the vacant dataset
calculate_local_morans <- function(data, variable, k = 5) {
  
  # Create spatial weights
  coords <- st_coordinates(st_centroid(data))
  neighbors <- knn2nb(knearneigh(coords, k = k))
  weights <- nb2listw(neighbors, style = "W", zero.policy = TRUE)
  
  # Calculate Local Moran's I
  local_moran <- localmoran(data[[variable]], weights)
  
  # Classify clusters
  mean_val <- mean(data[[variable]], na.rm = TRUE)
  
  data %>%
    mutate(
      local_i = local_moran[, 1],
      p_value = local_moran[, 5],
      is_significant = p_value < 0.05,
      
      moran_class = case_when(
        !is_significant ~ "Not Significant",
        local_i > 0 & .data[[variable]] > mean_val ~ "High-High",
        local_i > 0 & .data[[variable]] <= mean_val ~ "Low-Low",
        local_i < 0 & .data[[variable]] > mean_val ~ "High-Low",
        local_i < 0 & .data[[variable]] <= mean_val ~ "Low-High",
        TRUE ~ "Not Significant"
      )
    )
}

# Apply to vacant building dataset 
fishnet <- calculate_local_morans(fishnet, "countvacant", k = 5)
```

```{r visualize-morans}
#| fig-width: 8
#| fig-height: 6

# Visualize hot spots
ggplot() +
  geom_sf(
    data = fishnet, 
    aes(fill = moran_class), 
    color = NA
  ) +
  scale_fill_manual(
    values = c(
      "High-High" = "#d7191c",
      "High-Low" = "#fdae61",
      "Low-High" = "#abd9e9",
      "Low-Low" = "#2c7bb6",
      "Not Significant" = "gray90"
    ),
    name = "Cluster Type"
  ) +
  labs(
    title = "Local Moran's I: Vacant Buildings",
    subtitle = "High-High = Hot spots of disorder"
  ) +
  theme_crime()
```

```{r distance-to-hotspots}
# Get centroids of "High-High" cells (hot spots)
hotspots <- fishnet %>%
  filter(moran_class == "High-High") %>%
  st_centroid()

# Calculate distance from each cell to nearest hot spot
if (nrow(hotspots) > 0) {
  fishnet <- fishnet %>%
    mutate(
      dist_to_hotspot = as.numeric(
        st_distance(st_centroid(fishnet), hotspots %>% st_union())
      )
    )
  

  cat("  - Number of hot spot cells:", nrow(hotspots), "\n")
} else {
  fishnet <- fishnet %>%
    mutate(dist_to_hotspot = 0)
  cat("⚠ No significant hot spots found\n")
}
```

## Count Regression Models

```{r, results='hide'}
# Create clean modeling dataset

# load police districts and join them to fishnet
policeDistricts <- 
  st_read("https://data.cityofchicago.org/api/geospatial/24zt-jpfn?method=export&format=GeoJSON") %>%
  st_transform('ESRI:102271') %>%
  dplyr::select(District = dist_num)

# Join district information to fishnet
fishnet <- st_join(
  fishnet,
  policeDistricts,
  join = st_within,
  left = TRUE
) %>%
  filter(!is.na(District))  # Remove cells outside districts


cat("  - Districts:", length(unique(fishnet$District)), "\n")
cat("  - Cells:", nrow(fishnet), "\n")

fishnet_model <- fishnet %>%
  st_drop_geometry() %>%
  dplyr::select(
    uniqueID,
    kde_value,
    kde_b,
    local_i,
    countvacant,
    vacant.nn,
    dist_to_hotspot, 
    countBurglaries,
    District
  ) %>%
  na.omit()  # Remove any remaining NAs


districts <- unique(fishnet_model$District)
cv_results <- tibble()
```


```{r fit-poisson}
# Fit Poisson regression
model_poisson <- glm(
  countBurglaries ~  countvacant + vacant.nn + 
    dist_to_hotspot,
  data = fishnet_model,
  family = "poisson"
)

# Summary
summary(model_poisson)
```

```{r check-overdispersion}
# Calculate dispersion parameter
dispersion <- sum(residuals(model_poisson, type = "pearson")^2) / 
              model_poisson$df.residual

cat("Dispersion parameter:", round(dispersion, 2), "\n")
cat("Rule of thumb: >1.5 suggests overdispersion\n")

if (dispersion > 1.5) {
  cat("⚠ Overdispersion detected! Consider Negative Binomial model.\n")
} else {
  cat("✓ Dispersion looks okay for Poisson model.\n")
}
```
Negative Binomial

```{r fit-negbin}
# Fit Negative Binomial model
model_nb <- glm.nb(
  countBurglaries ~ countvacant + vacant.nn + 
    dist_to_hotspot,
  data = fishnet_model
)

# Summary
summary(model_nb)

# Compare AIC (lower is better)
cat("\nModel Comparison:\n")
cat("Poisson AIC:", round(AIC(model_poisson), 1), "\n")
cat("Negative Binomial AIC:", round(AIC(model_nb), 1), "\n")

```
## Spatial Cross-Validation (2017)

```{r spatial-cv}
cat("Running LOGO Cross-Validation...\n")

for (i in seq_along(districts)) {
  
  test_district <- districts[i]
  
  # Split data
  train_data <- fishnet_model %>% filter(District != test_district)
  test_data <- fishnet_model %>% filter(District == test_district)
  
  # Fit model on training data
  model_cv <- glm.nb(
    countBurglaries ~ countvacant + vacant.nn + 
      dist_to_hotspot,
    data = train_data
  )
  
  # Predict on test data
  test_data <- test_data %>%
    mutate(
      prediction = predict(model_cv, test_data, type = "response")
    )
  
  # Calculate metrics
  mae <- mean(abs(test_data$countBurglaries - test_data$prediction))
  rmse <- sqrt(mean((test_data$countBurglaries - test_data$prediction)^2))
  
  # Store results
  cv_results <- bind_rows(
    cv_results,
    tibble(
      fold = i,
      test_district = test_district,
      n_test = nrow(test_data),
      mae = mae,
      rmse = rmse
    )
  )
  
  cat("  Fold", i, "/", length(districts), "- District", test_district, 
      "- MAE:", round(mae, 2), "\n")
}

# Overall results
cat("Mean MAE:", round(mean(cv_results$mae), 2), "\n")
cat("Mean RMSE:", round(mean(cv_results$rmse), 2), "\n")
```


```{r cv-results-table}
# Show results
cv_results %>%
  arrange(desc(mae)) %>%
  kable(
    digits = 2,
    caption = "LOGO CV Results by District"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Model Evaluation


```{r final-predictions}
# Fit final model on all data
final_model <- glm.nb(
  countBurglaries ~ countvacant + vacant.nn + 
    dist_to_hotspot,
  data = fishnet_model
)

# Add predictions back to fishnet
fishnet <- fishnet %>%
  mutate(
    prediction_nb = predict(final_model, fishnet_model, type = "response")[match(uniqueID, fishnet_model$uniqueID)]
  )

# Also add KDE predictions (normalize to same scale as counts)
kde_sum <- sum(fishnet$kde_b, na.rm = TRUE)
count_sum <- sum(fishnet$countBurglaries, na.rm = TRUE)
fishnet <- fishnet %>%
  mutate(
    prediction_kde = (kde_b / kde_sum) * count_sum
  )
```

## Compare Model vs. KDE Baseline

```{r compare-models}
#| fig-width: 12
#| fig-height: 4

# Create three maps
p7 <- ggplot() +
  geom_sf(data = fishnet, aes(fill = countBurglaries), color = NA) +
  scale_fill_viridis_c(name = "Count", option = "plasma", limits = c(0, 15)) +
  labs(title = "Actual Burglaries") +
  theme_crime()

p8 <- ggplot() +
  geom_sf(data = fishnet, aes(fill = prediction_nb), color = NA) +
  scale_fill_viridis_c(name = "Predicted", option = "plasma", limits = c(0, 15)) +
  labs(title = "Model Predictions (Neg. Binomial)") +
  theme_crime()

p9 <- ggplot() +
  geom_sf(data = fishnet, aes(fill = prediction_kde), color = NA) +
  scale_fill_viridis_c(name = "Predicted", option = "plasma", limits = c(0, 15)) +
  labs(title = "KDE Baseline Predictions") +
  theme_crime()

p7 + p8 + p9 +
  plot_annotation(
    title = "Actual vs. Predicted Burglaries",
    subtitle = "Does our complex model outperform simple KDE?"
  )
```

```{r model-comparison-metrics}
# Calculate performance metrics
comparison <- fishnet %>%
  st_drop_geometry() %>%
  filter(!is.na(prediction_nb), !is.na(prediction_kde)) %>%
  summarize(
    model_mae = mean(abs(countBurglaries - prediction_nb)),
    model_rmse = sqrt(mean((countBurglaries - prediction_nb)^2)),
    kde_mae = mean(abs(countBurglaries - prediction_kde)),
    kde_rmse = sqrt(mean((countBurglaries - prediction_kde)^2))
  )

comparison %>%
  pivot_longer(everything(), names_to = "metric", values_to = "value") %>%
  separate(metric, into = c("approach", "metric"), sep = "_") %>%
  pivot_wider(names_from = metric, values_from = value) %>%
  kable(
    digits = 2,
    caption = "Model Performance Comparison"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r prediction-errors}
#| fig-width: 10
#| fig-height: 5

# Calculate errors
fishnet <- fishnet %>%
  mutate(
    error_nb = countBurglaries - prediction_nb,
    error_kde = countBurglaries - prediction_kde,
    abs_error_nb = abs(error_nb),
    abs_error_kde = abs(error_kde)
  )

# Map errors
p1 <- ggplot() +
  geom_sf(data = fishnet, aes(fill = error_nb), color = NA) +
  scale_fill_gradient2(
    name = "Error",
    low = "#2166ac", mid = "white", high = "#b2182b",
    midpoint = 0,
    limits = c(-10, 10)
  ) +
  labs(title = "Model Errors NB") +
  theme_crime()

p2 <- ggplot() +
  geom_sf(data = fishnet, aes(fill = abs_error_nb), color = NA) +
  scale_fill_viridis_c(name = "Abs. Error", option = "plasma") +
  labs(title = "Absolute Model Errors NB") +
  theme_crime()

p3 <- ggplot() +
  geom_sf(data = fishnet, aes(fill = error_kde), color = NA) +
  scale_fill_gradient2(
    name = "Error",
    low = "#2166ac", mid = "white", high = "#b2182b",
    midpoint = 0,
    limits = c(-10, 10)
  ) +
  labs(title = "Model Errors KDE") +
  theme_crime()

p4 <- ggplot() +
  geom_sf(data = fishnet, aes(fill = abs_error_kde), color = NA) +
  scale_fill_viridis_c(name = "Abs. Error", option = "plasma") +
  labs(title = "Absolute Model Errors KDE") +
  theme_crime()

p1 + p2 + p3 + p4
```


## Negative Binomial Model Table

```{r model-summary-table}
# Create nice summary table
model_summary <- broom::tidy(final_model, exponentiate = TRUE) %>%
  mutate(
    across(where(is.numeric), ~round(., 3))
  )

model_summary %>%
  kable(
    caption = "Final Negative Binomial Model Coefficients (Exponentiated)",
    col.names = c("Variable", "Rate Ratio", "Std. Error", "Z", "P-Value")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  footnote(
    general = "Rate ratios > 1 indicate positive association with burglary counts."
  )
```

## Written Analysis

**Intro**

The following passage will go through each major step to conduct this analysis in predicting burglaries in Chicago in 2017. 

This analysis integrated four types of datasets to perform its analysis: chicago boundaries, burglaries in 2017, 311 vacant building calls from 2017, and police districts. Vacant building calls were filtered to 2017 to align with the timeframe from the burglary dataset to avoid bias or enhance accuracy. Literature highlights that the presence of crime is associated with disorder in the built environment also known as "broken window theory", therefore 311 calls on vacant and abandoned buildings was utilized to understand if it can predict burglaries.

**Data Exploration**

Two maps are created to visualize the spatial distribution of vacant and abandoned buildings in Chicago. The point map displays the exact locations of reported buildings, while the density map (kernel density estimation) highlights areas of higher concentration.

Together, these visualizations provide an initial understanding of spatial patterns, revealing clusters and potential hotspots of vacancy across the city. This step is essential in exploratory data analysis, as it helps identify underlying spatial structure and informs subsequent modeling decisions, including the selection of relevant spatial features and predictors.

**Fishnet Grid**

In the next step, a fishnet grid of 500 by 500 meters is created to standardize the spatial unit of analysis. This grid allows for the aggregation of vacant building reports and burglary incidents within uniformly sized cells, making it possible to compare counts across space in a consistent way.

Aggregating the data to this grid helps reveal where both phenomena are concentrated and how frequently they occur within each area. However, this process comes with a tradeoff: while it improves comparability and enables spatial modeling, it does not preserve the exact point locations of individual events.

Additionally, the choice of grid size is methodologically important. Different grid resolutions can produce different spatial patterns, as larger cells may smooth variation while smaller cells may highlight more localized clustering. This reflects the Modifiable Areal Unit Problem (MAUP), where analytical results are sensitive to the scale and zoning of spatial units.

**Spatial Features**

A kernel density estimation (KDE) is developed in the next step to serve as a baseline non-parametric approach. This is implemented through point pattern analysis, where the spatial distribution of events is used to estimate a continuous surface of intensity. A bandwidth of 1000 meters is applied, allowing the model to capture the concentration of events within a defined spatial radius. The resulting KDE values are then extracted and assigned to the fishnet cells, enabling visualization of hotspots for both burglaries and vacant buildings.

Building on this, additional spatial features are constructed to incorporate spatial context into the regression model and assess whether patterns in vacant building reports can help predict burglary occurrences. The first feature is a k-nearest neighbor measure, calculated as the mean distance to the three nearest vacant buildings, which captures the degree of local clustering.


A distance-to-hotspot measure is created by identifying high-high clusters of vacant buildings and calculating the distance from each fishnet cell centroid to the nearest hotspot. This captures proximity to areas of concentrated vacancy, providing an additional spatial predictor for the model.

**Local Moran**

Next, a Local Moran’s I statistic is computed. Unlike Global Moran’s I, which summarizes overall spatial autocorrelation, Local Moran’s I identifies clustering at a more granular level, distinguishing between high-high clusters, low-low clusters, and spatial outliers. This function is applied to the fishnet using the count of vacant buildings, with k = 5 defining the number of nearest neighbors used to construct spatial weights. The outputs of this map indicate there are clusters of vacant buildings in Chicago, especially in central and southern areas. 

**Count Regression**

Now that spatial features have been created, they can be integrated into regression models to help understand if 311 vacant building calls can predict burglaries. Considering this is count data, a poisson model is utilized. The model utilizes count of vacant calls within each fishnet cell, the nearest neighbor spatial feature, and distance to hot spots to predict burglaries. The Poisson regression results indicate that vacant building reports are positively associated with burglary counts, specifically "countvacant".  Additionally, the mean distance to the three nearest vacant buildings has a small but statistically significant negative association, suggesting that burglaries are more likely to occur in areas where vacant buildings are more spatially clustered. In contrast, distance to vacant building hotspots is not statistically significant, indicating it does not meaningfully contribute to predicting burglary counts in this model. However, overdispersion was detected, with the model having a dispersion of 3.4 which is higher than the threshold of 1.5. This indicates that a negative binomial model is more suitable for this dataset. In the negative binomial model all three variables in the model were statistically significant and the mean distance to the three nearest vacant buildings still had a negative association with burglaries, while the remaining variables had positive associations. 

The negative binomial model shown at the end extracts coeffieicnts and trasforms them into interpretable results by changing them into rate ratios. Anything above 1 has a positive association with burgalries. Count vacant is the only variable that has a slight positive association, while the other variables were equal to 1 or less. 

The AIC is utilized to compare models to understand model performance. The negative binomial model performed better with an AIC of 7776.1 and the poisson model reporting an AIC of 9304.9.

**Validation**

A spatial cross validation was ran on police districits, and districts 3 and 12 are the hardest to forecast  due to their higher mse and rmse. This method is to understand how well our data can predict outcomes on unseen data. 

**Model Comparison**

Model comparison results indicate that the kernel density estimation (KDE) baseline outperforms the negative binomial model, with lower MAE (2.04 vs. 2.51) and RMSE (2.98 vs. 3.46). This suggests that burglary patterns are driven by underlying spatial clustering, which KDE captures effectively through smoothing. This highlights that our predictive variables are not comprehensible enough to predict burglaries, and other variables are necessary to create a model with stronger predictive power. 

However, while KDE provides stronger predictive accuracy, it does not offer insight into the relationship between vacant buildings and burglary. In contrast, the negative binomial model allows for interpretation of how vacant building prevalence and spatial features are associated with burglary counts.

This highlights an important tradeoff between predictive performance and interpretability: simpler spatial models may outperform in prediction, but the negative binomial provides coefficients for each variable allowing us to understand the magnitude of each variable's ability to predict the outcome, illustrating a more comprehensive picture of the variance within the dataset.

According to the maps, both models overpredict and underpreidct throughout the city. The negative binomial model overpredicts and underpredicts more than the KDE baseline. Additionally it performs poorly for areas along the southeastern parts of Chicago, the KDE baseline performs poorly for that area as well. There is a small pocket in the northern part of Chicago that the NB model does not accurately predict, this is also evident for the KDE baseline but not with the same magnitude. Broadly both models do not have large absolute errors, highlighting they both perform relatively well. 



